Inside Out: Externalizing Assumptions in Data Analysis as Validation Checks

### 2024-12-10 _manuscript/_tex/index.tex
test -> check

\usepackage{orcidlink}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

\begin{itemize}
\tightlist
\item
  test1: the test fails if the mean step count is below 8,200
\item
  test2: the test fails if the 30th percentile of the observed step
  counts is below 8,200
\item
  test3: the test fails if the standard deviation of the observed step
  counts exceeds 2,500.
\end{itemize}

The cutoff values chosen for these tests would presumably be chosen
based on prior experience with these kinds of data, but could also be
optimized using the method presented in the next section.

@@

While tests may score high on accuracy, they may be less effective at
explaining the various reasons behind unexpected results. This could
happen if, for example, a set of tests are all tangentially related to
the cause of the unexpected results, but none addresses the root cause.
It may also occur if the tests are highly correlated with one another,
leading to redundancy.

@@

A high mutual information value indicates redundancy among the tests,
while a low value suggests that the tests are independent and provides
unique information to diagnose the unexpected outcome. To standardize
this measure, the total correlation \emph{per observation} is
calculated, and an independence score, ranging between 0 and 1, is
defined as 1 - mutual information.

@@

Returning to the step count example introduced in Section~\ref{sec-toy},
the logic regression model is fitted to the three analysis checks
described previously to generate the prediction of the unexpected
outcome, which is whether the number of days with fewer than 8,000 steps
is greater than 5. The predictions from the logic regression model can
be compared with the simulated true outcome for calculating the
precision and recall metrics. Figure~\ref{fig-logic-reg} shows the
best-fitting logic regression model, which is a combination of test1
(``mean step count is below 8,200'') and test3 (``standard deviation
exceeds 2,500'') with an OR operator. In other words, if either test1 or
test3 is true, then we would predict an unexpected outcome in the
analysis (i.e.~the number of days with fewer than 8,000 steps is greater
than 5).

@@

\caption{\label{fig-logic-reg}Logic regression model fitted to the three
unit tests (test1, test2, test3). The model suggests using an OR rule to
combine test1 and test3 to predict the outcome expectation.}

@@

Table~\ref{tbl-logic-reg} presents the calculated precision, recall, and
independence for the three individual tests and the combined test rule
(test1 OR test3) from the logic regression. We also include the metric
calculated from fitting a regression tree model to the data to compare
the performance of the logic regression model. The harmonic and
arithmetic means are included to combine the three measures. The results
show that the two tests produced by the logic regression can accurately
predict 83.6\% cases of all \emph{actual unexpected results} in the
simulation data. Furthermore, 82.1\% of all \emph{predicted unexpected
results} were in fact observed to be unexpected.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2424}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1515}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1061}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1970}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1364}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}@{}}

\caption{\label{tbl-logic-reg}Accuracy (precision and recall) and
parsimony (independence) metrics for each individual unit test and for
the combined test rule (test1 OR test3) derived from the logic
regression model. The harmonic and arithmetic means of the three metrics
are included to evaluate the quality of the unit tests in diagnosing
unexpected step counts (more than five days with fewer than 8,000
steps).}

@@

\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
tests
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
precision
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
recall
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
independence
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
harmonic
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
arithmetic
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
test1 & 0.482 & 0.964 & 1.000 & 0.730 & 0.815 \\
test2 & 0.214 & 1.000 & 1.000 & 0.450 & 0.738 \\
test3 & 0.589 & 0.805 & 1.000 & 0.762 & 0.798 \\
test1 OR test3 & 0.821 & 0.836 & 0.999 & 0.879 & 0.886 \\
regression tree & 0.821 & 0.836 & 1.000 & 0.879 & 0.886 \\

@@

For comparison, the regression tree produces a similar prediction to the
logic regression, by first splitting on test1 and then splitting on
test3, and results in the same accuracy and overall score as the logic
regression model. However, we argue that the logic regression tree shown
in Figure~\ref{fig-logic-reg} is more interpretable for our purposes
because it provides a direct representation of which combinations of
analysis checks lead to unexpected outcomes. The logic regression tree
is also directly comparable to other diagnostic techniques, which we
discuss further in Section~\ref{sec-discussion}.

@@

The typical approach to studying the association between PM10 and
mortality is to apply a generalized linear model with a Poisson link to
relate daily mortality counts to daily measures of PM10. Based on
previous work and the range of effect sizes published in the literature,
an analyst might expect the coefficient for PM10 in this GLM to lie
between {[}0, 0.005{]}, after adjusting for daily temperature
\citep{samet200fine, welty2005acute}. Observing an estimated coefficient
outside of this interval would be highly unusual and would warrant a
serious re-examination of the analysis process. Therefore, our
unexpected outcome in this analysis will be the binary indicator of
whether the estimated coefficient for PM10 in the GLM lies outside of
the interval {[}0, 0.005{]}. In addition to providing a more substantial
problem for our methods, this example also demonstrates how the
procedure presented in Section~\ref{sec-method} can be used to select
cutoff values in the analysis checks to diagnose an unexpected PM10
coefficient from the generalized linear model.

@@

A logic regression is fitted using all variations of the tests in
Table~\ref{tbl-checks} to predict whether the PM10 coefficient is
unexpected. Figure~\ref{fig-linear-reg-tree} shows the optimal logic
regression tree from the fitted model. Precision, recall, and
independence score, along with their harmonic and arithmetic mean are
calculated in Table~\ref{tbl-linear-reg}.

@@

\caption{\label{fig-linear-reg-tree}Logic regression model fitted to the
fourteen unit tests and the outcome expectation (unexpected) as the
response variable. The model suggests the relationship: (sample size
\emph{greater than} 200 AND mortality-PM10 correlation \textless{}
-0.04) AND (mortality-temperature correlation \textgreater{} -0.35 OR
there exist mortality outlier) to predict the unexpected PM10
coefficient. {[}sorry the proper plot is not ready{]}.}

@@

\begin{longtable}[]{@{}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0882}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1471}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1029}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1765}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1912}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1324}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1618}}@{}}

\caption{\label{tbl-linear-reg}Accuracy (precision and recall) and
parsimony (independence) metrics derived from the logic regression
model, along with harmonic and arithmetic means, for individual unit
tests (1: sample size, 2: mortality-PM10 correlation, 3:
mortality-temperature correlation, 4: mortality outlier), and the
combined test rule 5: (sample size AND mortality-PM10 correlation) AND
(mortality-temperature correlation OR mortality outlier.}

@@

\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedleft
tests
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
precision
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
recall
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
overlapping
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
independence
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
harmonic
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
arithmetic
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
1 & 0.087 & 0.215 & 0 & 1 & 0.175 & 0.434 \\
2 & 0.988 & 0.610 & 0 & 1 & 0.822 & 0.866 \\
3 & 0.392 & 0.581 & 0 & 1 & 0.569 & 0.658 \\
4 & 0.649 & 0.641 & 0 & 1 & 0.732 & 0.763 \\
5 & 0.760 & 0.880 & 0 & 1 & 0.869 & 0.880 \\

@@

As shown in Figure~\ref{fig-linear-reg-tree}, there is no single
analysis check in the tree that predicts an unexpected outcome. rather
at least three checks in the tree must be TRUE in order for the model to
predict an unexpected outcome. Given the high independence of the tests
(Table~\ref{tbl-linear-reg}), this suggests that unexpected results are
only likely after multiple anomalies are observed in the data.

>>>>

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{orcidlink}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

\begin{itemize}
\tightlist
\item
  check 1: the check fails if the mean step count is below 8,200
\item
  check 2: the check fails if the 30th percentile of the observed step
  counts is below 8,200
\item
  check 3: the check fails if the standard deviation of the observed
  step counts exceeds 2,500.
\end{itemize}

The cutoff values chosen for these checks would presumably be chosen
based on prior experience with these kinds of data, but could also be
optimized using the method presented in the next section.

@@

While checks may score high on accuracy, they may be less effective at
explaining the various reasons behind unexpected results. This could
happen if, for example, a set of checks are all tangentially related to
the cause of the unexpected results, but none addresses the root cause.
It may also occur if the checks are highly correlated with one another,
leading to redundancy.

@@

A high mutual information value indicates redundancy among the checks,
while a low value suggests that the checks are independent and provides
unique information to diagnose the unexpected outcome. To standardize
this measure, the total correlation \emph{per observation} is
calculated, and an independence score, ranging between 0 and 1, is
defined as 1 - mutual information.

@@

Returning to the step count example introduced in Section~\ref{sec-toy},
the logic regression model is fitted to the three analysis checks
described previously to generate the prediction of the unexpected
outcome, which is whether the number of days with fewer than 8,000 steps
is greater than 5. The predictions from the logic regression model can
be compared with the simulated true outcome for calculating the
precision and recall metrics. Figure~\ref{fig-logic-reg} shows the
best-fitting logic regression model, which is a combination of check 1
(``mean step count is below 8,200'') and check 3 (``standard deviation
exceeds 2,500'') with an OR operator. In other words, if either check 1
or check 3 is true, then we would predict an unexpected outcome in the
analysis (i.e.~the number of days with fewer than 8,000 steps is greater
than 5).

@@

\caption{\label{fig-logic-reg}Logic regression model fitted to the three
unit tests (check 1, check 2, check 3). The model suggests using an OR
rule to combine check 1 and check 3 to predict the outcome expectation.}

@@

Table~\ref{tbl-logic-reg} presents the calculated precision, recall, and
independence for the three individual checks and the combined check rule
(check 1 OR check 3) from the logic regression. We also include the
metric calculated from fitting a regression tree model to the data to
compare the performance of the logic regression model. The harmonic and
arithmetic means are included to combine the three measures. The results
show that the two checks produced by the logic regression can accurately
predict 83.6\% cases of all \emph{actual unexpected results} in the
simulation data. Furthermore, 82.1\% of all \emph{predicted unexpected
results} were in fact observed to be unexpected.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.2468}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1299}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0909}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1688}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1169}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1429}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1039}}@{}}

\caption{\label{tbl-logic-reg}Accuracy (precision and recall) and
parsimony (independence) metrics for each individual check and for the
combined check rule (check 1 OR check 3) derived from the logic
regression model. The harmonic and arithmetic means of the three metrics
are included to evaluate the quality of the checks in diagnosing
unexpected step counts (more than five days with fewer than 8,000
steps).}

@@

\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
tests
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
precision
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
recall
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
independence
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
harmonic
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
arithmetic
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
checks
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
1 & 0.482 & 0.964 & 1.000 & 0.730 & 0.815 & check 1 \\
2 & 0.214 & 1.000 & 1.000 & 0.450 & 0.738 & check 2 \\
3 & 0.589 & 0.805 & 1.000 & 0.762 & 0.798 & check 3 \\
check 1 OR check 3 & 0.821 & 0.836 & 0.999 & 0.879 & 0.886 & NA \\
regression tree & 0.821 & 0.836 & 1.000 & 0.879 & 0.886 & NA \\

@@

For comparison, the regression tree produces a similar prediction to the
logic regression, by first splitting on check 1 and then splitting on
check 3, and results in the same accuracy and overall score as the logic
regression model. However, we argue that the logic regression tree shown
in Figure~\ref{fig-logic-reg} is more interpretable for our purposes
because it provides a direct representation of which combinations of
analysis checks lead to unexpected outcomes. The logic regression tree
is also directly comparable to other diagnostic techniques, which we
discuss further in Section~\ref{sec-discussion}.

@@

The typical approach to studying the association between PM10 and
mortality is to apply a generalized linear model with a Poisson link to
relate daily mortality counts to daily measures of PM10. Based on
previous work and the range of effect sizes published in the literature,
an analyst might expect the coefficient for PM10 in this GLM to lie
between \([0, 0.005]\), after adjusting for daily temperature
\citep{samet200fine, welty2005acute}. Observing an estimated coefficient
outside of this interval would be highly unusual and would warrant a
serious re-examination of the analysis process. Therefore, our
unexpected outcome in this analysis will be the binary indicator of
whether the estimated coefficient for PM10 in the GLM lies outside of
the interval \([0, 0.005]\). In addition to providing a more substantial
problem for our methods, this example also demonstrates how the
procedure presented in Section~\ref{sec-method} can be used to select
cutoff values in the analysis checks to diagnose an unexpected PM10
coefficient from the generalized linear model.

@@

A logic regression is fitted using all variations of the checks in
Table~\ref{tbl-checks} to predict whether the PM10 coefficient is
unexpected. Figure~\ref{fig-linear-reg-tree} shows the optimal logic
regression tree from the fitted model. Precision, recall, and
independence score, along with their harmonic and arithmetic mean are
calculated in Table~\ref{tbl-linear-reg}.

@@

\caption{\label{fig-linear-reg-tree}Logic regression model fitted to the
fourteen checks and the outcome expectation (unexpected) as the response
variable. The model suggests the relationship: (sample size
\emph{greater than} 200 AND mortality-PM10 correlation \textless{}
-0.04) AND (mortality-temperature correlation \textgreater{} -0.35 OR
there exist mortality outlier) to predict the unexpected PM10
coefficient. {[}sorry the proper plot is not ready{]}.}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.5368}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1053}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0737}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0737}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0947}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1158}}@{}}

\caption{\label{tbl-linear-reg}Accuracy (precision and recall) and
parsimony (independence) metrics derived from the logic regression
model, along with harmonic and arithmetic means, for individual checks
(1: sample size, 2: mortality-PM10 correlation, 3: mortality-temperature
correlation, 4: mortality outlier), and the combined check rule 5:
(sample size AND mortality-PM10 correlation) AND (mortality-temperature
correlation OR mortality outlier.}

@@

\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
check
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
precision
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
recall
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
indep.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
harmonic
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
arithmetic
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
check 1: sample size \textless= 200 & 0.087 & 0.215 & 1 & 0.175 &
0.434 \\
check 2: cor(m, PM10) \textless{} -0.03 & 0.988 & 0.610 & 1 & 0.822 &
0.866 \\
check 3: cor(m, tmp) \textgreater{} -0.35 & 0.392 & 0.581 & 1 & 0.569 &
0.658 \\
check 4: mortality\_outlier & 0.649 & 0.641 & 1 & 0.732 & 0.763 \\
(not check 1 AND check 2) AND (check 3 OR check 4) & 0.760 & 0.880 & 1 &
0.869 & 0.880 \\

@@

As shown in Figure~\ref{fig-linear-reg-tree}, there is no single
analysis check in the tree that predicts an unexpected outcome. rather
at least three checks in the tree must be TRUE in order for the model to
predict an unexpected outcome. Given the high independence of the checks
(Table~\ref{tbl-linear-reg}), this suggests that unexpected results are
only likely after multiple anomalies are observed in the data.
